---
title: "OUTLIER DETECTION IN TIMESERIES"
author: |
  | Amir Azarbakht
  | Michael Dumelle
  | Camden Lopez
  | Tadesse Zemicheal
fontsize: 12pt
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message = FALSE, 
  warning = FALSE, results = "hide", fig.height = 3, fig.width = 6) 

#Includes  
library('tsoutliers')
library('ggplot2')

```


## INTRODUCTION

## THEORETICAL RESULTS

### Types of outlier in ARIMA model 
We say that $x_t$ folows an ARIMA (p,d,q) model if $x_t$  has the form
\begin{align} \phi(B)(1-B)^dX_t = \theta(B)Z_t
\end{align}
Where B is backshift operator, $Z_t$ is white noise process, $\phi(B) = 1-\alpha_1B-\alpha_2B^2 ...-\alpha_pB^p$  $\theta(B) = 1+\beta_1B + \beta_2B^2 +... + \beta_pB^p$, and d is the number of differenced times. It is further assumed the root of $\phi(B)$ and $\theta(B)$ are assumed to be outside unit circle and have no common roots. 

#### Additive outliers (AO)
An additive outlier (AO) correspond to an exogenous change of a single observation of the time series. It is associated with inciden like measurement errors or impulse effect due to external effect.  A time series $y_1 ....y_T$ affected by AO at t=k is given by

\begin{equation} y_t = x_t + \omega I_t^k\end{equation}

```{r}
#simulation time series data 
set.seed(9)
# Arma model ARMA(1,1)
wn <- arima.sim(model=list(ar=c(0.5,0.3),ma=c(0.5,0.0),sd=1),200)
#add additive autlier with weight of 10 times to current value at two section 20 and 170
wn[20]<- 5*wn[20]
wn[170] <- 6*wn[170]
plot(1:200,wn,type='l',main="",xlab="Interval",ylab="") 
lines(18:21,wn[18:21],col='red')
lines(169:171,wn[169:171],col='red')
```

#### Innovative Outliers 

#### Temporal Change outliers 

#### Level Shifts Outliers

## OUTLIER DETECTION AND ESTIMATION PROCEDURE


## EXPERIMENTS 

## CONCULSION

## REFERENCE

## APPENDIX