---
title: "Outliers in Time Series"
subtitle: "ST 565 Final Project (Winter 2016)"
author: |
        | Amir Azarbakht, Michael Dumelle,
        | Camden Lopez & Tadesse Zemicheal
fontsize: 12pt
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
           results = "hide", fig.height = 3, fig.width = 6)
library('tsoutliers')
library('ggplot2')
```

# Introduction

Throughout the quarter, we have used various methods to model time series data.
However, we have not encountered any data that has been affected by the presence
of one or more outliers. In various statistical settings, outliers can have a
profound impact on the way data is analyzed. If outliers are not properly
adjusted for, inference on parameters and prediction of future observations
will be less reliable. In some extreme cases, it may be meaningless. Consider
a time series subject to the presence of outliers. How do we quantify an
outlier in the time series setting? Are the outliers seemingly random, or is
there some pattern among them? How do we properly account for the effect of
these outliers? These are all questions we intend to answer throughout this
report, and understanding them is crucial to correctly handling time series
data.

[*Goal is to model "normal" or "baseline" behavior without model fit being
influenced by anomalies (or measurement errors)*]

Throughout this report, we will look at four different types of outliers:
innovational outliers (IO), additive outliers (AO), temporary change outliers
(TC), and level shift outliers (LS). We intend to examine the effects that
these outliers have on standard ARIMA models. For simplicity, we will only
consider outliers in non-seasonal models. This problem becomes increasingly
complex when a seasonality component is added. We intend to use an iterative
algorithm described by Chen and Liu (1993) to detect these four outlier types.
This algorithm is implemented in the R package `tsoutliers`.
We will apply this algorithm to the `ipi` data set in `tsoutliers` and use
graphical tools to aid our analysis. This data set contains economic indices
from several European countries from 1999 to 2013. After successful detection,
we will also begin exploring ways to incorporate this information into our
model building process to obtain reliable parameter estimates and predictions.

# Outlier Models

We say that $X_t$ follows an ARIMA (p,d,q) model if it has the form

\begin{align}
X_t = \frac{\theta(B)}{\phi(B)(1-B)^d}Z_t 
\end{align}

where B is the standard backshift operator, $Z_t$ is white noise process
(identically and independently distributed as $N(0,\sigma^2)$), $\phi(B) =
1-\alpha_1B-\alpha_2B^2 ...-\alpha_pB^p$, $\theta(B) = 1- \beta_1B -
\beta_2B^2 -... - \beta_pB^q$, d is the order of differencing, and t = 1,...,n,
where n is the number of observations in the time series. It is further assumed
the roots of $\phi(B)$ and $\theta(B)$ are outside unit circle (this gives us
stationarity and invertibility) and have no common factors. Now, to describe
a time series which is influenced by a nonrepetitive event, we will consider
the following model

\begin{align}
X_t^* = X_t + \omega \frac{A(B)}{G(B)H(B)}I_1(t_1)
\end{align}

where $I_t(t_1)$ = 1 if t = $t_1$, and 0 otherwise. $I_t(t_1)$ acts as an
indicator function for the outlier impact occurence, with $t_1$ being the
location (possibly unknown) of the outlier. $\omega \frac{A(B)}{G(B)H(B)}$
denotes magnitude and pattern of some event. We will assume these are not
previously known. Outliers are then classified by imposing a structure on 
$\frac{A(B)}{G(B)H(B)}$ and estimating a value of $\omega$. In this report,
we will focus on four main types of outliers: Additive (AO), Innovational (IO),
Level Shift (LS), and Temporary Change (TC).

## Innovational Outlier (IO)

An innovational outlier is regarded as an initial impact with effects
lingering over future observations. For example, consider a time series
recording seismic wave activity in a specific area. Now, assume an earthquake
occurs. Seismic wave activity will increase dramatically, and the effects
of this earthquake will be present long time thereafter. More formally, we
define as innovational outlier as follows.

An innovational outlier at time t, is of the form 
\begin{align}
X_t^* = (X_t + \omega I_t(t_1))\frac{\phi(B)}{\alpha(B)\omega(B)}
\end{align}

One can see that IO's are obtained by letting $\frac{A(B)}{G(B)H(B)} =
\frac{\phi(B)}{\alpha(B)\omega(B)}$

It is important to note that these outliers are \textbf{not independent} of
the model, as they rely on $\theta(B)$, $\alpha(B)$, and $\phi(B)$ terms.
Thus, future values of the time series are affected by the IO. When an IO
outlier occurs at time $t = t_1$, the effect of this outlier on $Y_{t_1 + k} =
\omega \psi_k$, where $k \geq 0$ and $\psi_k$ is the $kth$ coefficient of the
polynomial $\psi_k$ is defined as 

\begin{align}
\psi(B) = \frac{\theta(B)}{\alpha(B)\phi(B)} = \psi_0 + \psi_1 B +
\psi_2 B^2 + .... + \psi_{max(p,d,q)}B^{max(p,q,d)}, \text{where} \ \psi_0 = 1
\end{align}

Because the time series is stationary and invertible, we have that the $\psi$'s
tend to 0. This implies that the IO we will deal with produce a temporary
effect (consider the sales example above) In other words, the IO at time $t_1$
does not effect the value in the time series at time $t_1 + k$, for large
enough $k$. In other cases, it is also possible that the IO produces an initial
effect and a subsequent permanent level shift.

## Additive Outlier (AO)

An additive outlier (AO) corresponds to an exogenous change of a single
observation of the time series. It is associated with incident like measurement
errors or impulse effect due to external forces. For an example of an additive
outlier, consider a time series recording highway traffic. If rainfall is
abnormally larger than expected, traffic would likely be much higher than
normal. More formally, an additive outlier at time t is defined as 

\begin{align}
X_t^* = X_t + \omega I_t(t_1)
\end{align}

```{r,echo=FALSE}
#simulation time series data 
set.seed(9)
# Arma model ARMA(1,1)
wn <- arima.sim(model=list(ar=c(0.5,0.3),ma=c(0.5,0.0),sd=1),200)
#add additive autlier with weight of 10 times to current value at two section 20 and 170
wn[20]<- 5*wn[20]
wn[170] <- 6*wn[170]
plot(1:200,wn,type='l',main="",xlab="Interval",ylab="") 
lines(18:21,wn[18:21],col='red')
lines(169:171,wn[169:171],col='red')
```

which is obtained by setting $\frac{A(B)}{G(B)H(B)} = 1$. The effect of these
outliers on the response at time t is \textbf{independent} of the ARIMA model
initially chosen, and does not depend on any function of B. We see that an
additive outlier simply acts as a shift in the value of the response when
$t = t_1$. The rest of the model remains unchanged when $t \neq t_1$, as
$I_t(t_1) = 0$.

## Level Shift (LS)

A level shift outlier (LS) produces a sudden, permanent change in the time
series values. Consider a time series of stock market prices for a given
company. Now, assume there is a new regulation the company must adhere to,
greatly changing how the company operates. Stock prices would likely jump
up or down and stay there, depending on the regulation. More formally, a
level shift outlier is defined as follows

\begin{align}
X_t^* = X_t + \frac{\omega I_t(t_1)}{(1 - B)}
\end{align}

which is obtained by setting $\frac{A(B)}{G(B)H(B)} = \frac{1}{(1 - B)}$.
Moreover, we see that these outliers are \textbf{independent} of the original
model, as they simply represent a shift in the response.

## Temporary Change (TC)

A temporary change (TC) outlier produces an abrupt change in the time series
which decays over time. For example, consider a time series recording daily
profit for a retail store. Now suppose there is some month long sporting sale
at the store which draws large crowds. The restaurant experiences inflated
sales due to the influx of people. It is likely that at the beginning of the
sale, a lot of people will visit the store. As the month progresses the total
sales per day should decrease, because many people have already gotten the
good deals. After the month ends, sales will go back to normal. This
illustrates the effect of the temporary change outlier. We rigorously define
the temporary change outlier as 

\begin{align}
X_t^* = X_t + \frac{\omega I_t(t_1)}{(1 - \delta B)}
\end{align}

We can see that $\frac{A(B)}{G(B)H(B)} = \frac{1}{(1 - \delta B)}$ and that
these outliers are \textbf{independent} of the model chosen.

```{r, fig.height=6, fig.width=6}
#Example of temporal shift outliers
#source http://stats.stackexchange.com/users/48766/javlacalle
tc <- rep(0, 50)
tc[20] <- 1
tc1 <- filter(tc, filter = 0, method = "recursive")
tc2 <- filter(tc, filter = 0.3, method = "recursive")
tc3 <- filter(tc, filter = 0.7, method = "recursive")
tc4 <- filter(tc, filter = 1, method = "recursive")
par(mfrow = c(2,2))
plot(tc1, main = "TC delta = 0")
plot(tc2, main = "TC delta = 0.3")
plot(tc3, main = "TC delta = 0.7")
plot(tc4, main = "TC delta = 1", type = "s")
#dev.off()
```

Notice that the temporary change outlier is a generalization of the additive
and level shift outliers ($\delta = 0$ and $\delta = 1$, respectively). In the
special level shift case, the time series actually does not decrease over time,
as it is a permanent change. $\delta$ is used to model the pace of the decaying
effect, and is often specified according to the specific needs of the
researcher.

# Outlier Detection and Estimation

In this section we consider the problem of detecting the presence of an
unknown number of outliers occuring at unknown times in a time series. The
outliers may be any of the four types described above. First, we describe
how the effects of a single outlier can be estimated using multiple regression.
Then we point out some practical issues, and the section culminates in a
description of the procedure proposed by Chen and Liu (1993) to jointly
estimate model parameters and outlier effects in a series.

## Estimating the Effect of a Single Outlier

When examining the effect of outliers, we must first assume the time series
parameters are known.  Next, define the polynomial

\begin{align}
\pi(B) = {\phi(B) \alpha (B) / \theta(B)} = 1 - \pi_1B - \pi_2B^2 - \pi_3B^3 - ...
\end{align}

where the $\pi_j$ weights for large $j$ become essentially 0, as the roots of
$\theta(B)$ all have absolute value greater than 1 (implying the infinite
series converges).  We can estimate residuals, $\hat{e_t}$, with the expression

\begin{align}
\hat{e_t} = \pi(B)Y_t^* \text{, for} \ t = 1,2,...
\end{align}

It is important to note that these residuals are subject to outlier effects.
We can express the residual for the four types of outliers by the formulas
below.

\begin{align}
\hat{e_t} &= \omega \pi(B) I_t(t_1) + a_t  & \text{Additive Outlier}\\
\hat{e_t} &= \omega I_t(t_1) + a_t & \text{Innovational Outlier}\\
\hat{e_t} &= \frac{\omega \pi(B) I_t(t_1)}{1 - B} + a_t & \text{Level Shift}\\
\hat{e_t} &= \frac{\omega \pi(B) I_t(t_1)}{1 - \delta B} + a_t & \text{Temporary Change}
\end{align}

We can express the above equations alternatively as $\hat{e_t} =
\omega x_{it} + a_t \text{, for} \ t = t_1, t_1 + 1, t_1 + 2, ... , n$ and
$i = 1,2,3,4,$ where

\begin{align}
&x_{it} = 0 \ \text{for all} \ i \ \text{and} \ t<t_1 \\
&x_{it_1} = 1 \ \text{for all} \ i \ \text{and} \ k \geq 1 \\
&x_{1(t_1 + k)} = 0 \\
&x_{2(t_1 + k)} = -\pi_k\\
&x_{3(t_1 + k)} = 1 - \sum_{j = 1}^{k} \pi_j\\
&x_{4(t_1 + k)} = \delta^{k - j} - \sum_{j = 1}^{k - 1} \delta^{k - j}\pi_j - \pi_k
\end{align}

Using the above formulas allows us obtain the least squares estimate for the
effect of an outlier, $\omega$, at $t = t_1$. These are provided below.

\begin{align}
& \hat{\omega}_{\text{AO}}(t_1) = \frac{\sum_{t = t_1}^{n} \hat{e}_tx_{2t}}{\sum_{t = t_1}^{n}x^2_{2t}}\\
& \hat{\omega}_{\text{IO}}(t_1) = \hat e_{t_1}\\
& \hat{\omega}_{\text{LS}}(t_1) = \frac{\sum_{t = t_1}^{n} \hat{e}_tx_{3t}}{\sum_{t = t_1}^{n}x^2_{3t}}\\
& \hat{\omega}_{\text{TC}}(t_1) = \frac{\sum_{t = t_1}^{n} \hat{e}_tx_{4t}}{\sum_{t = t_1}^{n}x^2_{4t}}\\
\end{align}

It is interesting to note that when $t_1 = n$ (last observation in a time
series) it is impossible to distinguish what type the outlier is, as $x_{it_1}
= 1$ for all $i$, and this is the only value in each sum term above.  Next,
one can standardize these estimated values to obtain asymptotically normal
test statistics of the outlier effects.

\begin{align}
& \hat{\tau}_{\text{AO}}(t_1) = \frac{\hat{\omega}_{\text{AO}}(t_1){\left(\sum_{t = t_1}^{n} x^2_{2t}\right)}^{1/2}}{\hat{\sigma}_a}\\
& \hat{\tau}_{\text{IO}}(t_1) = \frac{\hat{\omega}_{\text{IO}}(t_1)}{\hat{\sigma}_a}\\
& \hat{\tau}_{\text{LS}}(t_1) = \frac{\hat{\omega}_{\text{LS}}(t_1){\left(\sum_{t = t_1}^{n} x^2_{3t}\right)}^{1/2}}{\hat{\sigma}_a}\\
& \hat{\tau}_{\text{TC}}(t_1) = \frac{\hat{\omega}_{\text{TC}}(t_1){\left(\sum_{t = t_1}^{n} x^2_{4t}\right)}^{1/2}}{\hat{\sigma}_a}\\
\end{align}

where $\hat{\sigma}_a$ is an estimate of the standard deviation. We can
compare these test statistics to a standard normal distribution to find
appropriate p-values when testing $H_0: \omega = 0$ vs $H_1: \omega \neq 0$.

These results help us decide whether an unusual observation is actually an
outlier or is due to natural, random variation. Once we have deemed an
observation an outlier, it is possible to adjust a time series model to
account for this information and improve future predictions.

## Effects of Multiple Outliers

Estimation of outlier effects and calculation of statistics as described
above becomes complicated when multiple outliers are present. When it is assumed
(as above) that only one outlier is present, but in fact there are many, the
presence of the other outliers may bias the estimate of the effect of one
at the time $t$ under consideration. Or the effects of an outlier at $t$ may
be masked by the effects of an earlier outlier.

The problem of bias and masking suggests that the effects of all of the outliers
should be estimated jointly. But doing so when the number and time locations
of the outliers are unknown would require much more complex calculations.
The procedure described below detects outliers one at a time but attempts to
solve the bias and masking problem by repeatedly adjusting and re-examining
both the time series and the set of identified outliers.

## Estimating $\sigma_a$

The statistics described above require an estimate of $\sigma_a$. Since we
are assuming the presence of outliers which could bias the estimate upwards,
it is important to use an estimate of $\sigma_a$ that is robust to outliers.
Three options are the median absolute deviation (MAD), a trimmed standard
deviation, and the usual standard deviation of residuals calculated with the
residual at time $t_1$ omitted.

## Detection and Estimation Procedure

Chen and Liu (1993) proposed an iterative procedure for detecting outliers,
estimating their effects, and estimating the parameters of an ARIMA model
to fit the series after adjusting for outlier effects. The procedure assumes
that the order of the ARIMA model is known. We will see that implementation
of this procedure in `tsoutliers` can also automatically choose the order of
the ARIMA model.

The procedure has three stages. In Stage I, outliers are accumulated one-by-one
in order of descending magnitude of $\hat{\tau}$ statistic. In Stage II, any
outliers that can no longer be considered outliers after taking into account
the effects of all the others are dropped. In Stage III, the model estimates
are fixed at their final values and one last round of outlier accumulation-
elimination results in the final set of outliers and estimates of their effects.

### Stage I

1. Compute the maximum likelihood estimates of the model parameters.
2. For $t = 1, \dots, n$ calculate $\hat{\tau}_{IO}(t), \hat{\tau}_{AO}(t),
\hat{\tau}_{LS}(t),$ and $\hat{\tau}_{TC}(t)$. Let $\hat{\tau}_{max}$ be the
maximum of the absolute values of the $\hat{\tau}$ statistics.
If $\hat{\tau}_{max} > C$, where $C$ is a constant chosen beforehand,
there is potentially an outlier at the time and
of the type for which $\hat{\tau}_{max}$ occurred. Remove the effect of the
outlier from the series using equations above.
3. Repeat (2) until no more $\hat{\tau}$ statistics indicate potential outliers.
4. If any outliers have been found in (2--3), go back to (1), computing
the parameter estimates on the series that has now been adjusted for outlier
effects, and iterate through (2--3) again. If no outliers were found, proceed
to Stage II.

### Stage II

5. Jointly estimate the outlier effects using the same multiple regression
model above. Calculate the $\hat{\tau}$ statistics. If any of the statistics no
longer exceed the threshold $C$, drop the corresponding outlier.
6. Repeat (5) until no more outliers are dropped.
7. Obtain jointly estimated outlier effects using only the remaining outliers
and adjust the original series by removing their effects.
8. Compute the maximum likelihood estimates of the model parameters using
the adjusted series. If the change in residual standard error from the
previous estimate exceeds $\epsilon$, another threshold constant chosen
beforehand, go back to (5), using the new parameter estimates.
Otherwise, proceed to Stage III.

### Stage III

9. The parameter estimates found in (8) are the final estimates for the model.
Using only these model parameters, iterate through (2--3) and (5--6) to obtain
a final set of outliers and estimates of their effects.

## Comments

The procedure uses two threshold values, $C$ and $\epsilon$, which are chosen
by the user. Based on simulations, Chen and Liu (1993) recommend $C = 2.5$
to 2.9 for a shorter series of length less than $n = 100$ or so, and $C = 3.0$
for $n = 100$ to 200. For longer series, $n > 200$, $C$ greater than 3.0 may
be appropriate, but in any case multiple values should be tried to assess
how sensitive the results are to $C$. The value of $\epsilon$ controls the
accuracy of the parameter estimates, and a suggested value is 0.001.

# Examples 

### R package`tsoutliers` 
Following the approach described in Chen & Liu (1993), an automatic procedure for detection of outliers in time series is implemented in the package tsoutliers. This package implements a procedure based on the approach described in Chen and Liu (1993) for automatic detection of outliers in time series. Innovational outliers, additive outliers, level shifts,
temporary changes and seasonal level shifts are considered. 

The function `tso` is the main interface for the automatic procedure. The functions `locate.outliers.oloop`
and `remove.outliers` implement respectively the first and second stages of the procedure. In practice,
the user may stick to use the function `tso`. All the options at any stage of the procedure can be defined through the arguments passed to `tso`. (CRAN documentation)

`tso(y, xreg = NULL, cval = NULL, delta = 0.7, n.start = 50,
        types = c("AO", "LS", "TC"),
        maxit = 1, maxit.iloop = 4, cval.reduce = 0.14286,
        remove.method = c("en-masse", "bottom-up"),
        remove.cval = NULL,
        tsmethod = c("auto.arima", "arima", "stsm"),
        args.tsmethod = NULL, args.tsmodel = NULL, logfile = NULL)
`


[*Possible questions to address, in addition to showing one or two examples:*]

 * What functions does the `tsoutliers` package offer? What exactly do they do?
 In particular, how exactly is `auto.arima` used within `tso`?
 * What arguments need to be specified? What are the defaults? When should one
 provide arguments other than the defaults?
 * How does one decide which types of outliers to look for?
 * How does one choose $C$ and $\delta$?
 * What is the output of the R function(s) and how does one interpret it?

### Example 1:  Outlier in measurement of the annual flow of the river Nile at Ashwan from 1871-1970.

Fitting auto.arima model gives an ARIMA(1,1,1).

```{r,fig.align='left',echo=TRUE,eval=TRUE}
library(fma)

data(Nile)
plot(Nile)
fit<- auto.arima(Nile)
#fits with ARIMA(1,1,1)
print(fit)
```

Applying the two stage outlier detection process using `tsoutliers` function
generates

```{r,eval=TRUE,echo=TRUE}
nile.outliers <- tso(Nile,types=c("AO","LS","TC"))
plot(nile.outliers)
```
The outlier detected are 
```{r, echo=TRUE,results='markup'}
# Outlier values 
print(nile.outliers$outliers)
```
The main detected outliers are `level shift (LS)` in 1899 and `Additive outlier (AO)` in 1913.  Additinaly, the fitted ARIMA model is ARIMA( 0,0,0).

```{r, echo=TRUE,results='markup'}
# Outlier values 
print(nile.outliers$fit)
```


### Example 2. Outliers in price of chicken in USA from 1924 to 1993. The data shows the price was gradually decreasing, but there might be an outliers hidden in the trends. 
```{r,echo=TRUE}
library(fma)
plot(chicken,main="Chicken price in USA from 1924 to 1993")

```
Applying `'tsoutliers::tso` for 30 iteration, identifies the following outliers. 

```{r,echo=TRUE}
#check outliers for 30 maximum iteration
chicken.outlier <- tsoutliers::tso(chicken,types=c("AO","LS","TC","IO"),maxit.iloop=30)
```
The fitted ARIMA model  
```{r, echo=TRUE,results='markup'}
print(chicken.outlier$fit)
```
Detectd outliers
```{r, echo=TRUE,results='markup'}
# Outlier values 
plot(chicken.outlier)
#Summary of the detected outliers
print(chicken.outlier$outliers)
```
As we can see the main two outliers are `level shift outlier` in 1935. and `temporal change` outlier in 1943. 

# Conclusion

# References

Chang, I., Tiao, G. C., and Chen, C. (1988), "Estimation of Time Series
Parameters in the Presence of Outliers," *Technometrics*, 30, 193--204.

Chen, C. and Liu, Lon-Mu (1993), "Joint Estimation of Model Parameters and
Outlier Effects in Time Series," *Journal of the American Statistical
Association*, 88, 284--297.

Chen, C., and Tiao, G. C. (1990), "Random Level Shift Time Series Models,
ARIMA Approximation, and Level Shift Detection," *Journal of Business and
Economic Statistics*, 8, 170--186.

Fox, A. J. (1972), "Outliers in Time Series," *Journal of the Royal Statistical
Society*, Ser. B, 34, 350--363.

Tsay, Ruey S. (1986), "Time Series Model Specification in the Presence of
Outliers," *Journal of the American Statistical Association*, 81, 132--141.

Galeano,Pena (2013) "Finding Outliers in Linear and Nonlinear Time Series" Chapter
Robustness and Complex Data Structures pp 243-260

https://cran.r-project.org/web/packages/tsoutliers/tsoutliers.pdf

# Appendix

```{r, all-code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
